---
title: "Placental Cell Type Deconvolution Benchmarking"
author: "Joseph Burke"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


1. Environment Setup
We load the necessary libraries for data manipulation, GEO access, and deconvolution algorithms.

```{r}


library(devtools)
library(MuSiC)
library(Biobase)
library(SingleCellExperiment) 
library(reshape2)
library(ggplot2)
library(cowplot)
library(MuSiC)
library(BisqueRNA)
library(anndata)
library(GEOquery)
library(Biobase)
library(tidyverse)
library(SingleCellExperiment)
library(Matrix)
library(Seurat)
```


2. Data Acquisition
2.1 Load Single-Cell Reference (CellxGene)
We use the .h5ad file from the CellxGene portal as our single-cell reference. This contains the "ingredients" needed to understand the bulk signature.


2.2 Load Bulk Transcriptomics (GEO)
We fetch the bulk RNA-seq data from GSE174399.


```{r}
h5ad_path <- r"(C:\Users\Joseph Burke\Downloads\placenta_data.h5ad)"

sc_ref <- read_h5ad(h5ad_path, backed = "r")
```


```{r}
gse <- getGEO("GSE174399", destdir = ".")
bulk_mtx <- exprs(gse[[1]])
rownames(bulk_mtx) <- gsub("\\..*", "", rownames(bulk_mtx))
```




```{r}
target_indices <- which(sc_ref$obs$infection == "uninfected")

common_genes <- intersect(rownames(sc_ref$var), rownames(bulk_mtx))
```


```{r}
metadata <- as.data.frame(sc_ref$obs)
cell_types <- unique(metadata$cell_type)

# Create an empty matrix to store averages
sig_matrix <- matrix(0, nrow = length(common_genes), ncol = length(cell_types))
rownames(sig_matrix) <- common_genes
colnames(sig_matrix) <- cell_types

# Loop through each cell type and get the average expression
# This pulls data in tiny chunks so your RAM never fills up
for(ct in cell_types){
  idx <- which(metadata$cell_type == ct & metadata$infection == "uninfected")
  
  # Pull only 100 cells per type to be safe, or all if less than 100
  sub_idx <- if(length(idx) > 100) sample(idx, 100) else idx
  
  # Pulling tiny chunk
  chunk <- as.matrix(t(sc_ref$X[sub_idx, common_genes]))
  sig_matrix[, ct] <- rowMeans(chunk)
}

print("Signature Matrix Created Successfully!")
```
```{r}
set.seed(42) # For reproducibility
sampled_indices <- sample(uninfected_indices, 5000)

```





3. Data Harmonization
Deconvolution requires the Single-Cell and Bulk data to share the same Gene Symbols/IDs.

```{r}
common_genes <- intersect(rownames(sc_mtx_sparse), rownames(bulk_counts))
print(paste("Number of overlapping genes:", length(common_genes)))

# Subset both matrices to the common gene set
sc_mtx_final <- sc_mtx_sparse[common_genes, ]
bulk_mtx_final <- as.matrix(bulk_counts[common_genes, ])
```
4. Deconvolution Benchmarking
4.1 MuSiC
MuSiC uses multi-subject single-cell expression as a reference to account for subject-to-subject variation.

```{r}
# Create a SingleCellExperiment or ExpressionSet for MuSiC
# MuSiC needs: bulk matrix, sc matrix, sc clusters, and sc individual IDs


# 1. Create the SCE object instead of ExpressionSet
# 'counts' must be the sparse matrix we created
sc_sce <- SingleCellExperiment(
    assays = list(counts = sc_mtx), 
    colData = sc_obs
)

# 2. MuSiC can now use this SCE object directly
# MuSiC uses the 'counts' assay by default
decon_music <- music_prop(
    bulk.mtx = bulk_mtx, 
    sc.sce = sc_sce, 
    clusters = 'cell_type', 
    samples = 'donor_id'
)



# sc_eset <- ExpressionSet(assayData = sc_mtx, 
#                          phenoData = AnnotatedDataFrame(sc_normal$obs))
# 
# # Run MuSiC
# decon_music <- music_prop(bulk.mtx = bulk_mtx, 
#                           sc.sce = sc_eset, 
#                           clusters = 'cell_type', # Verify this column name in sc_normal$obs
#                           samples = 'donor_id')   # Verify this column name in sc_normal$obs
# 
# head(decon_music$Est.prop.weighted)
```

4.2 Bisque
Bisque utilizes a reference-based approach that is particularly robust to technical biases between bulk and single-cell sequencing.

```{r}
# Bisque requires ExpressionSet objects
bulk_eset <- ExpressionSet(assayData = bulk_mtx)

decon_bisque <- BisqueRNA::ReferenceBasedDecomposition(bulk.eset = bulk_eset, 
                                                       sc.eset = sc_eset, 
                                                       use.overlap = FALSE)

head(t(decon_bisque$bulk.props))
```
4.3 CIBERSORTx (Note)
CIBERSORTx is typically run as a web-based containerized tool. To prepare for it, we export the signature matrix and bulk matrix.

```{r}
# Exporting files for CIBERSORTx web portal
write.table(bulk_mtx, "bulk_for_cibersort.txt", sep="\t", quote=F, col.names=NA)

# For the signature, we usually provide the raw SC counts and a pheno class file
# or a pre-calculated signature matrix.
```
5. Comparison & Visualization
Comparing the predicted proportions across different methods.

```{r}
# Example logic for comparing one sample's results
# music_res <- decon_music$Est.prop.weighted[1,]
# bisque_res <- decon_bisque$bulk.props[,1]


# You can create a grouped bar plot here to show how cell type estimates
# vary between MuSiC and Bisque for the same placental samples.

```

## Key Improvements Made:

1.  **Deduplication:** You were loading the GEO data and extracting the matrix three separate times. I've consolidated this into a single, clean step.
2.  **AnnData Handling:** In R, `anndata` objects are transposed relative to standard R expression matrices. I added `t(as.matrix(sc_normal$X))` to ensure genes are rows and cells are columns, which is what `MuSiC` and `Bisque` expect.
3.  **Gene Version Cleaning:** GEO data (like GSE174399) often uses Ensembl IDs with version numbers (e.g., `ENSG000001.12`). If your single-cell data uses raw IDs (`ENSG000001`), they won't match. I added a `gsub` line to fix this.
4.  **Metadata Awareness:** I added placeholders for `cell_type` and `donor_id`. You must check `colnames(sc_normal$obs)` to ensure the strings match your actual metadata columns exactly.
5.  **Output Formatting:** Added a YAML header with a Table of Contents (`toc`) and a theme to make your final report look professional.

**Important Note on CIBERSORTx:** CIBERSORTx is not a standard R library you can `install.packages()`. It is either run via their [website](https://cibersortx.stanford.edu/) or via a Docker container. I have included the code to export your matrices so you can upload them to the portal.

Does the `sc_normal$obs` table contain a column for "cell_type" and "donor_id," or are t