---
title: "BMB410Project  Placental Cell Type Deconvolution Benchmarking"
author: "Joseph Burke"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


2. Environment Setup
We load the necessary libraries for data manipulation, GEO access, and deconvolution algorithms.


```{r Environment Setup}
library(anndata)              # Handling H5AD files [cite: 101]
library(Biobase)              # Base functions for Bioconductor [cite: 119]
library(BiocManager)          # Bioconductor package manager
library(BisqueRNA)            # Transformation-based alignment [cite: 327]
library(cowplot)              # Publication-ready ggplot2 themes
library(devtools)             # Software development tools
library(GEOquery)             # Bulk data retrieval [cite: 94]
library(ggplot2)              # Data visualization
library(Matrix)               # Sparse matrix handling [cite: 116]
library(MuSiC)                # Weight-based deconvolution [cite: 158]
library(reshape2)             # Data restructuring
library(scuttle)              # Single-cell analysis utilities
library(Seurat)               # QC, analysis, and exploration of SC data [cite: 106]
library(SingleCellExperiment) # Standardized SC container [cite: 98]
library(tidyverse)            # Data wrangling and visualization

# --- Environment & Memory Settings ---

# Set memory options for large placental atlas (approx. 158k cells) [cite: 91, 153]
options(future.globals.maxSize = 4000 * 1024^2) 

# Sort objects in your environment by size to monitor RAM usage 
sort(sapply(ls(), function(x) object.size(get(x))), decreasing = TRUE)
```



3.Data Preprocessing:Single-Cell: Transposing, downsampling, and metadata cleaning.  Bulk: Retrieving GEO data and harmonizing Gene IDs.
Harmonize the Datasets
I would focus on the "stay in your lane" rule for gene IDs. If the bulk data uses versioned Ensembl IDs (e.g., .12), they must be stripped to match the reference.

```{r}
# Load Bulk Data (GSE174399) [cite: 94]
gse <- getGEO("GSE174399", destdir = ".")
bulk_mtx <- exprs(gse[[1]])
# Clean version numbers to ensure intersection with SC reference 
rownames(bulk_mtx) <- gsub("\\..*", "", rownames(bulk_mtx))

# Load SC Reference (CELLxGENE Atlas) [cite: 90, 152]
sc_ref <- read_h5ad("placenta_data.h5ad")
# Essential: Transpose because AnnData is Cells x Genes, R expects Genes x Cells 
sc_mtx <- t(as.matrix(sc_ref$X))
```



4.Deconvolution Implementation: Separate sections for MuSiC, Bisque, and CIBERSORTx export.


Implement MuSiC (Weight-Based)
MuSiC is my choice for stability because it uses donor-to-donor variation to find the most "consistent" genes.

```{r}
# Create SCE object for MuSiC [cite: 170]
sc_sce <- SingleCellExperiment(assays = list(counts = sc_mtx), colData = sc_ref$obs)

# Run deconvolution
# Prioritize 'cell_type' and 'donor_id' to account for subject heterogeneity [cite: 170, 367]
decon_music <- music_prop(
    bulk.mtx = bulk_mtx, 
    sc.sce = sc_sce, 
    clusters = 'cell_type', 
    samples = 'donor_id'
)
```


Implement Bisque (Transformation-Based)
Bisque is the best choice for cross-platform alignment. I would implement this to see how it handles technical noise between the 10x reference and the bulk target.

```{r}
# Convert to ExpressionSet for Bisque [cite: 121, 352]
sc_eset <- ExpressionSet(assayData = sc_mtx, phenoData = AnnotatedDataFrame(as.data.frame(sc_ref$obs)))
bulk_eset <- ExpressionSet(assayData = bulk_mtx)

# Execute cross-platform bias alignment [cite: 328]
decon_bisque <- ReferenceBasedDecomposition(bulk.eset = bulk_eset, sc.eset = sc_eset)
```


5.Comparative Analysis: Statistical evaluation using Spearman Correlation ($\rho$) and RMSE. Statistical Assessment
Finally, I would benchmark the Compositional Accuracy by correlating these results with known placental biology—specifically looking for the expected >45% Trophoblast signal in villous tissue

```{r}
# Calculate Spearman Correlation for benchmarking [cite: 401]
# Compare MuSiC results to biological expectations [cite: 522]
cor_music <- cor(decon_music$Est.prop.weighted[, "Trophoblast"], 
                 expected_trends, method = "spearman")
```




6.Visualizations: Multi-panel plots comparing predicted proportions against biological expectations.




7.